---
title: "IDS investigation worksheet"
author: "by Knights Templar: Tudor,Mengchuan Sun, Jiacheng Zhang,Binxi Yang"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(lubridate)
```


```{r load-data}
# load your data 
formula_e <- read.csv("data/formula_e_race_results.csv")
formula_e <- formula_e %>% mutate(race_city = substring(race_name,6))
formula_e %>% mutate(race_date = ymd(race_date)) %>% arrange(race_date)
```
```{r choose rank 1 data}
formula_e_rank1 <- filter(formula_e, rank==1) %>%
  select(., season, race_city, time_retired) %>% arrange(season) %>%
  mutate(time_retired = case_when(nchar(time_retired) == 9 ~ paste0("0:", time_retired),
                                  nchar(time_retired) > 9 ~ time_retired)) %>%
  filter(is.na(time_retired) == FALSE)
  
formula_e_rank1 %>%
  count(race_city)

formula_e_rank1 %>%
  arrange(time_retired)

formula_e_rank1_mini <- formula_e_rank1 %>%
  filter(race_city == "Berlin" | 
           race_city == "Marrakesh" |
           race_city == "Mexico City" |
           race_city == "Paris")

formula_e_rank1_mini <- formula_e_rank1_mini %>%
  mutate(time_retired = as.numeric(seconds(hms(time_retired))))

formula_e_rank1_mini
```

```{r tidy retired_time}
formula_e_tidy<- formula_e %>% 
  filter(rank != "DSQ") %>%
  filter(rank != "NC") %>%
  filter(nchar(time_retired) >= 4) %>% 
  filter(grepl("a", time_retired) == FALSE) %>% 
  filter(str_detect(time_retired, "^[[:alpha:]]") == FALSE) %>% 
  mutate(race_date = ymd(race_date)) %>% 
  arrange(race_date)

formula_e_tidy$id <- c(1:nrow(formula_e_tidy))

formula_e_tidy[311,13] <- "56:27.535"
formula_e_tidy[323,13] <- "+1:01.365"
formula_e_tidy[324,13] <- "+1:09.646"
formula_e_tidy[471,13] <- "+39.137"
formula_e_tidy[472,13] <- "+55.189"
formula_e_tidy[548,13] <- "+1:13.880"
formula_e_tidy[549,13] <- "+1:31.382"
formula_e_tidy[662,13] <- "47:02.511"
formula_e_tidy[835,13] <- "+1:27.00912"
```

```{r create object for time calculation}
formula_e_time <- select(formula_e_tidy, race_date, id, time_retired) %>% 
  mutate(tidying_time = gsub("s", "", time_retired, fixed=TRUE)) %>% 
  mutate(finish_time = gsub(" ", "", tidying_time)) %>% 
  mutate(list_of_time = strsplit(gsub(",", "", finish_time), ":")) %>% 
  mutate(finish_hour = 0) %>% 
  mutate(finish_minute = 0) %>% 
  mutate(finish_second = 0.0) %>% 
  mutate(second_behind = 0.0) %>%
  mutate(total_second = 0.0) %>% 
  arrange(race_date)
```

```{r calculation}
i <- 1
while (i <= nrow(formula_e_time)){
  formula_e_time[i,7] <- case_when(
    length(formula_e_time[i,6][[1]]) == 3 ~ as.numeric(formula_e_time[i,6][[1]][1]), 
    TRUE ~ 0)
  formula_e_time[i,8] <- case_when(
    length(formula_e_time[i,6][[1]]) == 3 ~ as.numeric(formula_e_time[i,6][[1]][2]), 
    length(formula_e_time[i,6][[1]]) == 2 ~ as.numeric(formula_e_time[i,6][[1]][1]), 
    TRUE ~ 0)
  formula_e_time[i,9] <- case_when(
    length(formula_e_time[i,6][[1]]) == 3 ~ as.double(formula_e_time[i,6][[1]][3]), 
    length(formula_e_time[i,6][[1]]) == 2 ~ as.double(formula_e_time[i,6][[1]][2]), 
    length(formula_e_time[i,6][[1]]) == 1 ~ as.double(formula_e_time[i,6][[1]][1]))
  i <- i+1
}

bufferdate <- 0
buffer_hour <- 0
buffer_minute <- 0
buffer_second <- 0
i <- 1
while (i <= nrow(formula_e_time)){
  if (bufferdate != formula_e_time[i,1]){
    bufferdate <- formula_e_time[i,1] 
    buffer_hour <- formula_e_time[i,7] 
    buffer_minute <- formula_e_time[i,8] 
    buffer_second <- formula_e_time[i,9] 
    formula_e_time[i,11] <- 3600*buffer_hour + 60*buffer_minute + buffer_second
  }else{
    formula_e_time[i,10] <- 3600*formula_e_time[i,7] + 60*formula_e_time[i,8] + formula_e_time[i,9]
    formula_e_time[i,11] <- 3600*(buffer_hour + formula_e_time[i,7]) + 60*(buffer_minute + formula_e_time[i,8]) + buffer_second + formula_e_time[i,9]
  }
  i <- i+1
}
```


```{r joining result to formula_e_tidy}
formula_e_time <- select(formula_e_time, id, second_behind, total_second)
formula_e_tidy = inner_join(formula_e_tidy, formula_e_time, by = "id")
```

```{r}
a <- formula_e_tidy %>%
  group_by(race_date) %>% 
  summarise(avg_sec_b = mean(second_behind), 
            med_sec_b = median(second_behind), 
            sd_sec_b = sd(second_behind), 
            min_sec = min(total_second), 
            avg_sec = mean(total_second), 
            med_sec = median(total_second), 
            max_sec = max(total_second))

race_statistics <- count(formula_e_tidy, race_date) 
race_statistics <- left_join(race_statistics, a, by = "race_date")
```

```{r investigations about time distribution}
ggplot(race_statistics, 
       aes(x = race_date, 
           y = avg_sec, 
           colour = avg_sec_b)) + 
  geom_point() + 
  labs(x = "date", 
       y = "mean time to finish the race (in second)", 
       colour = "mean time other drivers
left behind (in second)")

ggplot(race_statistics, 
       aes(x = avg_sec, 
           y = sd_sec_b, 
           colour = avg_sec_b)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = "y ~ x") + 
  labs(x = "mean time to finish the race (in second)", 
       y = "standard difference of time cost", 
       colour = "mean time other drivers
left behind (in second)")

ggplot(race_statistics, 
       aes(x = avg_sec_b, 
           y = sd_sec_b)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = "y ~ x") + 
  labs(x = "mean time other drivers left behind (in second)", 
       y = "standard difference of time cost")

ggplot(race_statistics, 
       aes(x = race_date, 
           y = sd_sec_b, 
           colour = avg_sec_b)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = "y~x") + 
  labs(x = "date", 
       y = "standard difference of time cost", 
       colour = "mean time other drivers
left behind (in second)")
```

```{r calculate average rank}
formu <- formula_e %>% 
  filter(is.numeric(as.numeric(rank))) %>% 
  mutate(nrank=as.numeric(rank)) %>% 
  select(team, nrank) %>% 
  group_by(team) %>% 
  summarise(vAverage_Team_rank=mean(nrank, na.rm = TRUE))
formula_e <- left_join(formula_e, formu, by = "team")

formu <- formula_e %>% 
  filter(is.numeric(as.numeric(rank))) %>% 
  mutate(nrank=as.numeric(rank)) %>% 
  select(driver, nrank) %>% 
  group_by(driver) %>% 
  summarise(Average_rank=mean(nrank, na.rm = TRUE))
formula_e <- left_join(formula_e, formu, by = "driver")
```
```{r rank 1 times over time}

formula_e_rank1_mini %>%
  ggplot(mapping = aes(x = season, y = time_retired, color = race_city, group_by=race_city)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm", formula='y~x')

```
##Tudor: Finish time vs season graph & fix time format
##Sun: Frequency table of teams reaching top 5 ranks in each race
##Lucien: Average time distribution
##ybx: Modelling

```{r top 5 frequency}

formula_e_with_nrank <- formula_e %>%
  filter(is.numeric(as.numeric(rank))) %>% 
  mutate(nrank = as.numeric(rank)) 

top_5_teams <- formula_e_with_nrank %>%
  filter(nrank <= 5) %>%
  select(season, race_city, team, nrank) %>%
  arrange(season, race_city, nrank)

top_5_teams_summary <- top_5_teams %>%
  group_by(season, race_city, team) %>%
      mutate(team = case_when(team == "Audi Sport ABT" ~ "Audi",
                            team == "e.dams-Renault" ~ "e.Dams-Renault",
                            TRUE ~ team)) %>%
  count()

print(top_5_teams_summary)

selected_cities <- c("Mexico City", "Berlin")#Berlin is more staight

selected_cities_data <- top_5_teams_summary %>%
  filter(race_city %in% selected_cities)

top_teams_selected_cities <- selected_cities_data %>%
  group_by(team) %>%
  summarise(total_frequency = sum(n), .groups = "keep") %>%
  top_n(3, total_frequency)

filtered_data <- selected_cities_data %>%
  filter(team %in% top_teams_selected_cities$team)

histogram_top_teams_selected_cities <- ggplot(filtered_data, aes(x = race_city, fill = factor(team))) +
  geom_bar(position = "dodge", stat = "count") +
  labs(title = "Top Teams' Performance in Selected Cities",
       x = "Race Track",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_discrete(name = "Team")

print(histogram_top_teams_selected_cities)

mean_frequency_by_season <- top_5_teams_summary %>%
  group_by(season) %>%
  summarise(mean_frequency = mean(n), .groups = "keep")

line_plot_mean_frequency <- ggplot(mean_frequency_by_season, aes(x = season, y = mean_frequency)) +
  geom_line() +
  geom_point() +
  labs(title = "Mean Frequency of Top 5 Finishes by Season",
       x = "Season",
       y = "Mean Frequency") +
  theme_minimal()

print(line_plot_mean_frequency)

top5_by_city_and_season <- formula_e %>%
  filter(rank %in% 1:5,race_city %in% c("Mexico City", "Berlin")) %>%
  group_by(season,race_city,team) %>%
  summarise(total_top5_count = n(), .groups = "keep") %>%
  arrange(season,race_city,team)

sum_of_seasons <- formula_e %>%
  filter(rank %in% 1:5, race_city %in% c("Mexico City", "Berlin")) %>% 
  group_by(race_city,team) %>%
  summarise(total_top5_count = sum(rank %in% 1:5), .groups = "keep")

  print(top5_by_city_and_season)

  print(sum_of_seasons)
  #Berlin - 6, mexico - 5

  
  top_5_teams_summary <- top_5_teams_summary %>%
    filter( race_city %in% c("Mexico City", "Berlin"))
  
  top_5_mean_totals <- top_5_teams_summary %>%
    group_by(race_city, team) %>%
    summarise(total = sum(n)) %>%
    mutate(mean_total = case_when(race_city == "Berlin" ~ total / 6,
                                  race_city == "Mexico City" ~ total / 5)) %>%
    filter(mean_total > 0.2)


  top_5_mean_totals %>%
    ggplot(mapping = aes(x = team, y = mean_total, fill = team)) +
    geom_bar(stat = "identity")+
    facet_grid(cols = vars(race_city)) + 
    coord_flip() 
    

  print(top_5_mean_totals) 
 ## print(histogram_top_5_mean_totals)
 



```



