---
title: "IDS investigation worksheet"
author: "by Knights Templar: Tudor,Mengchuan Sun, Jiacheng Zhang,Binxi Yang"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(lubridate)



```


```{r load-data}
# load your data 
formula_e <- read.csv("data/formula_e_race_results.csv")
formula_e <- formula_e %>% mutate(race_city = substring(race_name,6))
formula_e %>% mutate(race_date = ymd(race_date)) %>% arrange(race_date)

formula_e_rank1 <- filter(formula_e, rank==1) %>%
  select(., season, race_city, time_retired) %>% arrange(season) %>%
  mutate(time_retired = case_when(nchar(time_retired) == 9 ~ paste0("0:", time_retired),
                                  nchar(time_retired) > 9 ~ time_retired)) %>%
  filter(is.na(time_retired) == FALSE)
  
formula_e_rank1 %>%
  count(race_city)

formula_e_rank1 %>%
  arrange(time_retired)

formula_e_rank1_mini <- formula_e_rank1 %>%
  filter(race_city == "Berlin" | 
           race_city == "Marrakesh" |
           race_city == "Mexico City" |
           race_city == "Paris")

formula_e_rank1_mini
```


```{r calculate average rank}
formu <- formula_e %>% 
  filter(is.numeric(as.numeric(rank))) %>% 
  mutate(nrank=as.numeric(rank)) %>% 
  select(team, nrank) %>% 
  group_by(team) %>% 
  summarise(vAverage_Team_rank=mean(nrank, na.rm = TRUE))
formula_e <- left_join(formula_e, formu, by = "team")

formu <- formula_e %>% 
  filter(is.numeric(as.numeric(rank))) %>% 
  mutate(nrank=as.numeric(rank)) %>% 
  select(driver, nrank) %>% 
  group_by(driver) %>% 
  summarise(Average_rank=mean(nrank, na.rm = TRUE))
formula_e <- left_join(formula_e, formu, by = "driver")
```
```{r rank 1 times over time}

formula_e_rank1_mini %>%
  ggplot(mapping = aes(x = season, y = time_retired, color = race_city, group_by=race_city)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm", formula='y~x')

```
##Tudor: Finish time vs season graph & fix time format
##Sun: Frequency table of teams reaching top 5 ranks in each race
##Lucien: Average time distribution
##ybx: Modelling

```{r top 5 frequency}

formula_e_with_nrank <- formula_e %>%
  filter(is.numeric(as.numeric(rank))) %>% 
  mutate(nrank = as.numeric(rank)) 

top_5_teams <- formula_e_with_nrank %>%
  filter(nrank <= 5) %>%
  select(season, race_city, team, nrank) %>%
  arrange(season, race_city, nrank)

top_5_teams_summary <- top_5_teams %>%
  group_by(season, race_city, team) %>%
  count()

print(top_5_teams_summary)

selected_cities <- c("Mexico City", "Riyadh", "Monaco", "Paris")

selected_cities_data <- top_5_teams_summary %>%
  filter(race_city %in% selected_cities)

top_teams_selected_cities <- selected_cities_data %>%
  group_by(team) %>%
  summarise(total_frequency = sum(n), .groups = "keep") %>%
  top_n(3, total_frequency)

filtered_data <- selected_cities_data %>%
  filter(team %in% top_teams_selected_cities$team)

histogram_top_teams_selected_cities <- ggplot(filtered_data, aes(x = race_city, fill = factor(team))) +
  geom_bar(position = "dodge", stat = "count") +
  labs(title = "Top Teams' Performance in Selected Cities",
       x = "Race Track",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_discrete(name = "Team")

print(histogram_top_teams_selected_cities)

```



