---
title: "An Analysis of factors that affect <br> race completion times and probability <br> of winning in Formula E"
author: "Error 404 <br> Tudor Cucu, Binxi Yang, Jiacheng Zhang, Mengchuan Sun"
institute: "University of Edinburgh"
date: "29 Nov 2023"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      
---

```{r install-package, include = FALSE, eval = FALSE}
# Copy and paste the following code into your console to download and install
# the `xaringan` package that contains all of the code which allows you 
# to create presentation slides in Rmarkdown
```


```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r load-data, include=FALSE}
# Load your data here
formula_e <- read.csv("data/formula_e_race_results.csv")
formula_e <- formula_e %>% mutate(race_city = substring(race_name,6))
formula_e %>% mutate(race_date = ymd(race_date)) %>% arrange(race_date)
```

```{r include=FALSE}

#Background image
style_xaringan(
  title_slide_text_color = "black", 
  title_slide_background_color = "#fAfAfA",
  title_slide_background_image = "img/formulaE_logopng.png",
  title_slide_background_size = "20%",
  title_slide_background_position = "7% 98%"
)
```

class: middle

- How do average race completion times vary over time?
- What factors affect the rank achieved by race drivers?
- To predict which teams are more likely to win certain races.
- To gain insights into how some teams build their cars.
---

class: inverse, center, middle

# Data Source

---

# Formula E Championship dataset $^*$

.pull-left[
The main variables of interest for our investigation: 
]
.pull-right[
- time_retired (detailing the time taken to complete the race) <br>
- team <br>
- race_name (including the city in which the race took place as well as
the year) <br>
- rank
]

.footnote[
[*] Reference at the last slide
]
---

class: inverse, middle, center

# Investigate

---

# tidying data

- The city in which each race was held is mutated as race_city in order to get a variable represents different race tracks.

- A few cities which had the largest number of races is selected to prevent effects of track difference.

- A graph showcasing the evolution of rank 1 times over the seasons in selected cities is created.

- Different lines are plotted correspond to each city.

---

class: inverse, center, middle

# Prediction

---
class: middle

# Our expectations

We assume that rank 1 times would go down over the seasons <br> as the drivers gain more experience with the tracks and as the technology in the cars advances.

---

# Graph of selected cities

```{r choose rank 1 data, message=FALSE, warning=FALSE, echo=FALSE}
formula_e_rank1 <- filter(formula_e, rank==1) %>%
  select(., season, race_city, time_retired) %>% arrange(season) %>%
  mutate(time_retired = case_when(nchar(time_retired) == 9 ~ paste0("0:", time_retired),
                                  nchar(time_retired) > 9 ~ time_retired)) %>%
  filter(is.na(time_retired) == FALSE)
  

formula_e_rank1_mini <- formula_e_rank1 %>%
  filter(race_city == "Berlin" | 
           race_city == "Marrakesh" |
           race_city == "Mexico City" |
           race_city == "Paris")

formula_e_rank1_mini <- formula_e_rank1_mini %>%
  mutate(time_retired = as.numeric(seconds(hms(time_retired))))



formula_e_rank1_mini %>%
  ggplot(mapping = aes(x = season, y = time_retired, color = race_city, group_by=race_city)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm", formula='y~x') +
  labs(title = "Evolution of Rank 1 times over time in selected cities",
       x = "Season",
       y = "Time (s)",
       color = "Race City")
  
```

---

# Tidy "time_retired"

- In order to obtain some more conclusive results, we decided to utilise all of the race times.

.pull-left[
- haven't find a function to change the format of "+(mm:)ss.sss" directly <br> 
- original data not in uniform format
]
.pull-right[
e.g. <br> 
56:27.535  -> 56:27:535 <br> 
+1:27.0092 -> +1:27.009,2 <br>
47:02.511  -> 47'02.511
]

```{r tidy retired_time, echo=FALSE}
formula_e_tidy<- formula_e %>% 
  filter(rank != "DSQ") %>%
  filter(rank != "NC") %>%
  filter(nchar(time_retired) >= 4) %>% 
  filter(grepl("a", time_retired) == FALSE) %>% 
  filter(str_detect(time_retired, "^[[:alpha:]]") == FALSE) %>% 
  mutate(race_date = ymd(race_date)) %>% 
  arrange(race_date)

formula_e_tidy$id <- c(1:nrow(formula_e_tidy))

formula_e_tidy[311,13] <- "56:27.535"
formula_e_tidy[323,13] <- "+1:01.365"
formula_e_tidy[324,13] <- "+1:09.646"
formula_e_tidy[471,13] <- "+39.137"
formula_e_tidy[472,13] <- "+55.189"
formula_e_tidy[548,13] <- "+1:13.880"
formula_e_tidy[549,13] <- "+1:31.382"
formula_e_tidy[662,13] <- "47:02.511"
formula_e_tidy[835,13] <- "+1:27.00912"
```
```{r create object for time calculation, echo=FALSE}
formula_e_time <- select(formula_e_tidy, race_date, id, time_retired) %>% 
  mutate(tidying_time = gsub("s", "", time_retired, fixed=TRUE)) %>% 
  mutate(finish_time = gsub(" ", "", tidying_time)) %>% 
  mutate(list_of_time = strsplit(gsub(",", "", finish_time), ":")) %>% 
  mutate(finish_hour = 0) %>% 
  mutate(finish_minute = 0) %>% 
  mutate(finish_second = 0.0) %>% 
  mutate(second_behind = 0.0) %>%
  mutate(total_second = 0.0) %>% 
  arrange(race_date)
```
```{r calculation, echo=FALSE}
i <- 1
while (i <= nrow(formula_e_time)){
  formula_e_time[i,7] <- case_when(
    length(formula_e_time[i,6][[1]]) == 3 ~ as.numeric(formula_e_time[i,6][[1]][1]), 
    TRUE ~ 0)
  formula_e_time[i,8] <- case_when(
    length(formula_e_time[i,6][[1]]) == 3 ~ as.numeric(formula_e_time[i,6][[1]][2]), 
    length(formula_e_time[i,6][[1]]) == 2 ~ as.numeric(formula_e_time[i,6][[1]][1]), 
    TRUE ~ 0)
  formula_e_time[i,9] <- case_when(
    length(formula_e_time[i,6][[1]]) == 3 ~ as.double(formula_e_time[i,6][[1]][3]), 
    length(formula_e_time[i,6][[1]]) == 2 ~ as.double(formula_e_time[i,6][[1]][2]), 
    length(formula_e_time[i,6][[1]]) == 1 ~ as.double(formula_e_time[i,6][[1]][1]))
  i <- i+1
}

bufferdate <- 0
buffer_hour <- 0
buffer_minute <- 0
buffer_second <- 0
i <- 1
while (i <= nrow(formula_e_time)){
  if (bufferdate != formula_e_time[i,1]){
    bufferdate <- formula_e_time[i,1] 
    buffer_hour <- formula_e_time[i,7] 
    buffer_minute <- formula_e_time[i,8] 
    buffer_second <- formula_e_time[i,9] 
    formula_e_time[i,11] <- 3600*buffer_hour + 60*buffer_minute + buffer_second
  }else{
    formula_e_time[i,10] <- 3600*formula_e_time[i,7] + 60*formula_e_time[i,8] + formula_e_time[i,9]
    formula_e_time[i,11] <- 3600*(buffer_hour + formula_e_time[i,7]) + 60*(buffer_minute + formula_e_time[i,8]) + buffer_second + formula_e_time[i,9]
  }
  i <- i+1
}
```
```{r joining result to formula_e_tidy, echo=FALSE}
formula_e_time <- select(formula_e_time, id, second_behind, total_second)
formula_e_tidy = inner_join(formula_e_tidy, formula_e_time, by = "id")
```
```{r, echo=FALSE}
a <- formula_e_tidy %>%
  group_by(race_date) %>% 
  summarise(avg_sec_b = mean(second_behind), 
            med_sec_b = median(second_behind), 
            sd_sec_b = sd(second_behind), 
            min_sec = min(total_second), 
            avg_sec = mean(total_second), 
            med_sec = median(total_second), 
            max_sec = max(total_second))

race_statistics <- count(formula_e_tidy, race_date) 
race_statistics <- left_join(race_statistics, a, by = "race_date")
```
---

```{r investigations about time distribution, echo =FALSE, warning=FALSE, out.height = "75%"}
ggplot(race_statistics, 
       aes(x = race_date, 
           y = avg_sec, 
           colour = avg_sec_b)) + 
  geom_point() + 
  geom_smooth(method = "glm", formula = "y ~ x") +
  labs(title = "Mean time to finish the race vs. Date the race held", 
       x = "Date", 
       y = "Mean Time to Finish the Race (s)", 
       colour = "Mean Time Other Drivers
Left Behind (s)")
```
---

# Are Cars Different?

We wanted to conduct research in was whether different teams build their cars differently and how this would affect their probability of achieving a high rank on certain tracks.

---
# Our Prediction

If a team performs better on tracks with many long straight lines than on tracks with many turns, then it may suggest that this team builds cars with high top speed and acceleration in a straight line, but poorer maneuverability

---

```{r top 5 frequency, echo=FALSE, warning=FALSE, message=FALSE}

formula_e_with_nrank <- formula_e %>%
  filter(is.numeric(as.numeric(rank))) %>% 
  mutate(nrank = as.numeric(rank)) 

top_5_teams <- formula_e_with_nrank %>%
  filter(nrank <= 5) %>%
  select(season, race_city, team, nrank) %>%
  arrange(season, race_city, nrank)

top_5_teams_summary <- top_5_teams %>%
  group_by(season, race_city, team) %>%
      mutate(team = case_when(team == "Audi Sport ABT" ~ "Audi",
                            team == "e.dams-Renault" ~ "e.Dams-Renault",
                            TRUE ~ team)) %>%
  count()

#print(top_5_teams_summary)

selected_cities <- c("Mexico City", "Berlin")#Berlin is more staight

selected_cities_data <- top_5_teams_summary %>%
  filter(race_city %in% selected_cities)

top_teams_selected_cities <- selected_cities_data %>%
  group_by(team) %>%
  summarise(total_frequency = sum(n), .groups = "keep") %>%
  top_n(3, total_frequency)

filtered_data <- selected_cities_data %>%
  filter(team %in% top_teams_selected_cities$team)

histogram_top_teams_selected_cities <- ggplot(filtered_data, aes(x = race_city, fill = factor(team))) +
  geom_bar(position = "dodge", stat = "count") +
  labs(title = "Top Teams' Performance in Selected Cities",
       x = "Race Track",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_discrete(name = "Team")

#print(histogram_top_teams_selected_cities)

mean_frequency_by_season <- top_5_teams_summary %>%
  group_by(season) %>%
  summarise(mean_frequency = mean(n), .groups = "keep")

line_plot_mean_frequency <- ggplot(mean_frequency_by_season, aes(x = season, y = mean_frequency)) +
  geom_line() +
  geom_point() +
  labs(title = "Mean Frequency of Top 5 Finishes by Season",
       x = "Season",
       y = "Mean Frequency") +
  theme_minimal()

#print(line_plot_mean_frequency)

top5_by_city_and_season <- formula_e %>%
  filter(rank %in% 1:5,race_city %in% c("Mexico City", "Berlin")) %>%
  group_by(season,race_city,team) %>%
  summarise(total_top5_count = n(), .groups = "keep") %>%
  arrange(season,race_city,team)

sum_of_seasons <- formula_e %>%
  filter(rank %in% 1:5, race_city %in% c("Mexico City", "Berlin")) %>% 
  group_by(race_city,team) %>%
  summarise(total_top5_count = sum(rank %in% 1:5), .groups = "keep")

  #print(top5_by_city_and_season)

  #print(sum_of_seasons)
  #Berlin - 6, mexico - 5

  
  top_5_teams_summary <- top_5_teams_summary %>%
    filter( race_city %in% c("Mexico City", "Berlin"))
  
  top_5_mean_totals <- top_5_teams_summary %>%
    group_by(race_city, team) %>%
    summarise(total = sum(n)) %>%
    mutate(mean_total = case_when(race_city == "Berlin" ~ total / 6,
                                  race_city == "Mexico City" ~ total / 5)) %>%
    filter(mean_total > 0.2)

#print(top_5_mean_totals) 
  
  top_5_mean_totals %>%
    ggplot(mapping = aes(x = team, y = mean_total, fill = team)) +
    geom_bar(stat = "identity")+
    facet_grid(cols = vars(race_city)) + 
    coord_flip() +
    labs(
         x = "Proportion of races where top 5 rank was achieved",
         y = "Team") + 
    guides(fill = "none")
    

```

Berlin track: 10 corners <br> 
Mexico track: 16 corners
---

# Model predicting "time_retired" variable

```{r calculate average rank, include=FALSE}
formu <- formula_e %>% 
  filter(is.numeric(as.numeric(rank))) %>% 
  mutate(nrank=as.numeric(rank)) %>% 
  select(team, nrank) %>% 
  group_by(team) %>% 
  summarise(vAverage_Team_rank=mean(nrank, na.rm = TRUE))
formula_e <- left_join(formula_e, formu, by = "team")

formu <- formula_e %>% 
  filter(is.numeric(as.numeric(rank))) %>% 
  mutate(nrank=as.numeric(rank)) %>% 
  select(driver, nrank) %>% 
  group_by(driver) %>% 
  summarise(Average_rank=mean(nrank, na.rm = TRUE))
formula_e <- left_join(formula_e, formu, by = "driver")
```

```{r Prepare the data required for modeling, message=FALSE, warning=FALSE, include=FALSE}
glimpse(formula_e)

data <- formula_e %>% 
  filter(is.numeric(as.numeric(rank))) %>%
  mutate(date = ymd(race_date),
         time_retired = as.numeric(seconds(hms(time_retired))),
         rank = as.numeric(rank),
         city = case_when(
           grepl("Berlin race",race_city) ~ "Berlin",
           grepl("Diriyah",race_city) ~ "Diriyah",
           grepl("Hong Kong",race_city) ~ "Hong Kong",
           grepl("London",race_city) ~ "London",
           grepl("New York",race_city) ~ "New York",
           grepl("Montreal",race_city) ~ "Montreal",
           TRUE ~ race_city),
         is_top5 = case_when(
           rank <= 5 ~ "1",
           TRUE ~ "0"),
         is_top5 = fct_relevel(is_top5,"1","0"),
         vaverage_team_rank = vAverage_Team_rank,
         average_rank = Average_rank
         ) %>% 
  select(season,race_num,date,team_group,rank,time_retired,points,city,is_top5,vaverage_team_rank,average_rank)

data <- drop_na(data)
knitr::kable(head(data,10))

set.seed(2023)
data_split <- initial_split(data)
train_data <- training(data_split)
test_data  <- testing(data_split)
```

```{r linear model, message=FALSE, warning=FALSE, echo=FALSE, out.width = "55%", out.height = "80%"}
formula_e_linear_rec <- recipe(time_retired ~ ., data = data) %>%
  step_dummy(all_nominal(), -all_outcomes())

formula_e_linear_mod <- linear_reg() %>% 
  set_engine("lm")

formula_e_linear_wflow <- workflow() %>% 
  add_model(formula_e_linear_mod) %>% 
  add_recipe(formula_e_linear_rec)

formula_e_linear_fit <- formula_e_linear_wflow %>% 
  fit(data = train_data)

#tidy(formula_e_linear_fit)

formula_e_linear_pred <- predict(formula_e_linear_fit, test_data) %>% 
  bind_cols(test_data) 

#formula_e_linear_pred

num <- 1:length(formula_e_linear_pred$.pred)
ggplot(formula_e_linear_pred) + 
  geom_line(aes(x = num, y = .pred),color = "red")+
  geom_line(aes(x = num, y = time_retired),color = "green")+
  labs(title = "Predictions of linear models on the test_data",
       x = NULL,
       y = "Time Retired")
```

```{r echo = FALSE}
glance(formula_e_linear_fit)$adj.r.squared
```
---

# Model predicting probability of reaching top 5

.pull-left[
```{r logistic model, message=FALSE, warning=FALSE, echo=FALSE, out.width = "120%"}
formula_e_logistic_rec <- recipe(is_top5 ~ ., data = data) %>%
  step_rm(rank,date,points,city) %>% 
  step_dummy(all_nominal(), -all_outcomes())

formula_e_logistic_mod <- logistic_reg() %>% 
  set_engine("glm")

formula_e_logistic_wflow <- workflow() %>% 
  add_model(formula_e_logistic_mod) %>% 
  add_recipe(formula_e_logistic_rec)

formula_e_logistic_fit <- formula_e_logistic_wflow %>% 
  fit(data = train_data)

#knitr::kable(tidy(formula_e_logistic_fit))

formula_e_logistic_pred <- predict(formula_e_logistic_fit, test_data, type = "prob") %>% 
  bind_cols(test_data) 

#formula_e_logistic_pred

formula_e_logistic_pred %>%
  roc_curve(
    truth = is_top5,
    .pred_0,
    event_level = "second"
  ) %>%
  autoplot()
```
]
.pull-right[
```{r echo = FALSE}
formula_e_logistic_pred %>%
  roc_auc(
    truth = is_top5,
    .pred_0,
    event_level = "second"
  )

```
]
---

# Reference

Dataset: 

MLandry, 2020, Formula E Championship, Electronic Dataset, Kaggle.com, <https://www.kaggle.com/datasets/mlandry/formula-e-championship>, accessed 2023.10.20

References: 

Wikipedia, 2023, List of Formula E ePrix, Wikipedia, <https://en.wikipedia.org/wiki/List_of_Formula_E_ePrix>, accessed at 2023.111.26

title image:

WCPSeries, 2023, [FORMULA E] SEASON 2016/17, <https://images.app.goo.gl/tjoG3zAtxxHeBNX69>, accessed at 2023.11.30